DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-03-07 manuscript/manuscript.Rmd
workflow notes

abstract: |
  Date estimates for times of evolutionary diveregences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication and policy decisions.

 Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms

  When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
  Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.

@@

  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.

@@

1. It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree. If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths.
1. Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard. To do this, `datelife` implements the Taxonomic Name Resolution Service (TNRS; @Boyle2013) provided by OpenTree [@opentreeAPIs]. TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
1. The current version of `datelife` only accepts scientific taxonomic names as input. Names can belong to any taxonomic group or binomial specific. If an input taxon name belongs to an "inclusive" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, `datelife` retrieves all species names within the "inclusive" taxonomic group and adds them to the input. If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
1. The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided. The \texttt{datelifeQuery} object is used next to search the chronogram database.
1. Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips. Then, each pruned chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
1. At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
1. To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree. Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user. Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum andmaximum ages. Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
1. If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
1. Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.

>>>>

abstract: |
  Date estimates for times of evolutionary diveregences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication and policy decisions.
  Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
  When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
  Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.

@@

  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.

@@

1. It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths.
1. Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard. To do this, `datelife` implements the Taxonomic Name Resolution Service (TNRS; @Boyle2013) provided by OpenTree [@opentreeAPIs]. TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
1. The current version of `datelife` only accepts scientific taxonomic names as input. Names can belong to any taxonomic group or binomial specific. If an input taxon name belongs to an "inclusive" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, `datelife` retrieves all species names within the "inclusive" taxonomic group and adds them to the input. If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
1. The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided. The \texttt{datelifeQuery} object is used next to search the chronogram database.
1. Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips. Then, each pruned chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
1. At this point, summary data is collected to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, ??? is this only for ones that have all the tax?  c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
1. To construct summary trees we use a fixed topology, either provided by the user, taken from the literature <!--The user would have to take it from the literature, ya?--> or obtained from expert phylogenetic information, such as the OpenTree synthetic tree. 
1. ????? More meat on the methods needed: e.g. how the distances are generated and summarized, how conflict is handled...
1. Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user. Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum and maximum ages. Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
1. If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how? 
1. Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.

### 2022-03-07 manuscript/manuscript.tex
workflow notes

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3452}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

\abstract{
Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology. Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.

@@

The main workflow of \texttt{datelife} is to construct a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life (OpenTree), which incorporates phylogenetic data from the TreeBASE database as well.
We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms. Additionally, time data extracted from source chronograms can be usedas secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods.

Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology. How well this chronograms work for this purpose still needs to be tested.

\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
}

@@

  It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree. If input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004). The input tree is not required to have branch lengths.

@@

  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.

@@

  To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree. Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user. Summarizing method options available include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and any summary statistics such as median, minimum andmaximum ages. Finally, \texttt{datelife} applies the summarized time distances as node calibrations to date the given topology with BLADJ (Campbell O. Webb \& Donoghue, 2005).

@@

  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.

>>>>

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3484}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

\abstract{
Date estimates for times of evolutionary diveregences are key data for research in the natural sciences. These estimates also provide valuable information for for education, science communication and policy decisions.
Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.

@@

The main workflow of \texttt{datelife} is to construct a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life (OpenTree), which incorporates phylogenetic data from the TreeBASE database as well.
We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms. Additionally, time data extracted from source chronograms can be used as secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods external to datelife.

\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
}

@@

  It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004). The input tree is not required to have branch lengths.

@@

  At this point, summary data is collected to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, ??? is this only for ones that have all the tax? c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.

@@

  To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree.

@@

  ????? More meat on the methods needed: e.g.~how the distances are generated and summarized, how conflict is handled\ldots{}
  Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user. Summarizing method options available include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and any summary statistics such as median, minimum and maximum ages. Finally, \texttt{datelife} applies the summarized time distances as node calibrations to date the given topology with BLADJ (Campbell O. Webb \& Donoghue, 2005).
  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how?
